Induced universal graphs for families of small graphs

### 2021-10-19 paper/paper.tex
More writing

We write $\calF(k)$ to denote the family of all graphs on $k$ vertices,
and we write $f(k)$ to denote the order (that is, the
number of vertices) of a minimal induced universal graph for $\calF(k)$.
We write $F(k)$ to denote the number of non-isomorphic graphs of order $f(k)$
that are induced universal for $f(k)$.
To give an example, we have $f(3)=5$ and $F(3)=5$. All five of the minimal induced universal graphs
for $\calF(3)$ are shown in \Cref{fig:univ3a}. \Cref{fig:univ3b} shows the four
graphs in $\calF(3)$ as induced subgraphs of a single induced universal graph.

@@

    \par\bigskip
    \subfigure[][A demonstration that the first graph in \Cref{fig:univ3a} is induced universal
            for $\calF(3)$.  For each graph $G$ in $\calF(3)$ ($I_3$, $K_3$, $P_3$, and
            a graph with a single edge), an induced subgraph isomorphic to $G$ is shown in a
            single color.] {
      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                clockwise=5] {
                    0,1[fill=uofgpumpkin],2[fill=uofgpumpkin],3[fill=uofgpumpkin],4;
           3 -- 0;
           4 -- 0;
           4 -- 1;
           4 -- 3;
        };
      \end{tikzpicture}
      \qquad
      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                clockwise=5] {
                    0[fill=uofgturquoise],1,2,3[fill=uofgturquoise],4[fill=uofgturquoise];
           4 -- 1;
           3 --[uofgturquoise, ultra thick] 0;
           4 --[uofgturquoise, ultra thick] 0;
           4 --[uofgturquoise, ultra thick] 3;
        };
      \end{tikzpicture}
      \qquad
      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                clockwise=5] {
                    0,1[fill=uofgrose],2,3[fill=uofgrose],4[fill=uofgrose];
           3 -- 0;
           4 -- 0;
           4 --[uofgrose, ultra thick] 1;
           4 --[uofgrose, ultra thick] 3;
        };
      \end{tikzpicture}
      \qquad
      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                clockwise=5] {
                    0,1[fill=uofgmocha],2[fill=uofgmocha],3,4[fill=uofgmocha];
           3 -- 0;
           4 -- 0;
           4 --[uofgmocha, ultra thick] 1;
           4 -- 3;
        };
      \end{tikzpicture}
      \label{fig:univ3b}
    }
\caption{Induced universal graphs for the family of all graphs on 3 vertices}
\label{fig:graphs}
\end{figure}

This paper uses a brute-force approach similar to that of Preen to find minimal
induced universal graphs for families of small \emph{general} graphs.
Section~\ref{sec:method} describes our computer search method, including an
optimisation to reduce run times.  Sections~\ref{sec:results5} presents exact
values of $f(k)$ for $k \leq 5$ along with
the corresponding counts of minimal induced universal subgraphs.
Section~\ref{sec:f6} gives the value of $f(6)$ and and
Section~\ref{sec:f7} gives bounds on $f(7)$; in each of these sections, the lower
bound is proven and a graph obtained by heuristic search demonstrating the
upper bound is given.  Finally, Section~\ref{sec:trees} gives results for
families of trees.

@@

Given $\calF$ and $n$,
we use the brute-force approach described by Preen \cite{preen_math_se} and shown
in \Cref{alg:brute-force} to find the set of
$n$-vertex graphs that are induced universal for the family $\calF$.
The function $\FuncSty{AllInducedUniversalGraphs}()$ is the entry point.
\Lineref{TryEachG} tests in turn each candidate graph $G$ from the family
of $n$-vertex graphs, and adds to the collection $\calG$ those that are induced universal
for $\calF$.  The function $\FuncSty{IsInducedUniversal}()$ tests
whether a graph $G$ is induced universal for $\calF$ by checking that
$H$ is isomorphic to a subgraph of $G$ for each $G \in \calF$.

For the collection $\calF(n)$ of candidate graphs on \lineref{TryEachG}, our
implementation uses trees generated using Brendan McKay's \texttt{geng}
program \cite{DBLP:journals/jal/McKay98}
\footnote{The program \texttt{geng} is distributed
with Nauty; we used version 27.r3.\url{https://pallini.di.uniroma1.it/}}.
These are read in \texttt{graph6}
format\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/formats.html}}
from a text file as the program proceeds, and therefore only one candidate
graph at a time needs to be stored in memory.

For the call $\FuncSty{InducedSubIso}()$ to an algorithm for the induced
subgraph isomorphism decision problem on line \lineref{CallSubIso},
our program uses a Python implementation of the McSplit
algorithm \cite{DBLP:conf/ijcai/McCreeshPT17}.
McSplit was designed for
the more general maximum common induced subgraph problem,
but it can be used for induced subgraph isomorphism with a simple
modification: we backtrack when the calculated upper bound
on the order of a common subgraph is less the order of graph $H$.
This method for solving
the subgraph isomorphism problem is very fast if both input graphs
are small, as the graphs considered in this paper are.
While McSplit is suitable for our purposes in this paper,
we do not claim that it is the the fastest induced subgraph isomorphism
solver for very small graphs; it would be useful future work to perform
an experimental comparison with other subgraph isomorphism solvers.
%%% (Although
%%% I have described the application of this method to families of all graphs of a given order,
%%% the method could easily be applied to any family $\calF$ of graphs, and indeed
%%% Section~\ref{sec:trees} gives results from the application of this method
%%% to families of trees.)

\begin{algorithm}[h!]
\DontPrintSemicolon
\nl $\FuncSty{IsInducedUniversal}(\calF,G)$ \label{BruteForceFun} \;
\nl \KwData{A family of graphs $\calF$ and a graph $G$.}
\nl \KwResult{A logical value indicating whether $G$ is induced universal for $\calF$.}
\nl \Begin{
\nl    \For {$H \in \calF$ \label{HLoop}}{
\nl      \lIf {$\neg\FuncSty{InducedSubIso}(H, G)$\label{CallSubIso}}{$\KwSty{return}$~$\AlgVar{false}$}
       }
\nl   $\KwSty{return}$~$\AlgVar{true}$ \;
    }
\medskip
\nl $\FuncSty{AllInducedUniversalGraphs}(\calF,n)$ \label{BruteForceFun} \;
\nl \KwData{A family of graphs $\calF$ and a natural number $n$.}
\nl \KwResult{The set of all graphs of order $n$ that are induced universal for $\calF$.}
\nl \Begin{
\nl    $\calG \gets \emptyset$ \;
\nl    \For {$G \in \calF(n)$}{
\nl      \lIf {$\FuncSty{IsInducedUniversal}(\calF,G)$\label{TryEachG}}{$\calG \gets \calG \cup \{G\}$}
       }
\nl   $\KwSty{return}$~$\calG$ \;
    }
\caption{A brute-force algorithm for finding all order-$n$ induced universal subgraphs of a family $\calF$ of graphs.
	The entry point is $\FuncSty{AllInducedUniversalGraphs}(\calF,n)$.}
\label{alg:brute-force}
\end{algorithm}

The full set of experiments described in this paper, run sequentially,
took less than an hour to complete on a laptop with an Intel Core i5-6200U CPU
and 8 GB RAM.\footnote{The code and
results from this paper, including lists of minimal induced universal graphs
in graph6 format, are available from
\url{https://github.com/jamestrimble/small-universal-graphs}}

@@

When deciding on \lineref{HLoop} of the algorithm whether a graph $G$ contains
every element of $\calF$ as an induced subgraph, are some iteration orders of
$\calF$ better than others?  For a graph $G$ that contains a copy of each
element of $\calF$, the iteration order is of no importance; we much check
every element of $\calF$ before returning $\AlgVar{true}$.  For a graph $G$ that
does \emph{not} contain a copy of each element of $\calF$, however, we would like
to iterate over $\calF$ in an order such that graphs that are likely to fail the
subgraph isomorphism test are checked early; that way, we can quickly return
$\AlgVar{false}$.

We begin by considering the case where $\calF = \calF(k)$ for some $k$; that is,
$\calF$ is the set of all graphs of order $k$.
As Diaconis and Chatterjee note, \cite{chatterjee2021isomorphisms}, the complete
graph $K_n$ is the `hardest' graph of order $n$ for a random graph to contain.
This led me
to consider two possible orderings in which $K_k$ and its complement $I_k$ come
first.  The first of these approaches is to the sort $\calF(k)$ such that
graphs with large automorphism groups appear first.  (Intuitively, if a graph
has few automorphisms then we can generate many different labelled graphs
by permuting the vertex labels.  This gives more `opportunities' for the
graph to be an induced subgraph of $G$.)
The second approach considered is to place graphs
with unusually high or low edge counts, as measured by the absolute value of
${2|E(G)| - {|V(G)| \choose 2}}$, first.

\Cref{fig:scatter} examines whether graphs that appear early in the orderings
of $\calF$
generated by each of these two strategies are contained in few graphs with
a given number of vertices, as we would hope to be the case.
For this experiment, we let $\calF = \calF(5)$, and we consider the problem
of finding induced universal graphs of order 8.
Each point on the plots
represents one of the 34 graphs of order 5 (fewer than 34 points are visible
due to exactly-overlapping points, as we would expect).  On the vertical
axis, we have the number of graphs in $\calF(8)$ containing $G$; we would like
graphs for which this value is small to appear early in our order.
On the horizontal axes we have each of our two measures: size of automorphism group
and `extremeness' of edge count.  The first of these measures has a strong correlation
with the number of graphs containing $G$ as an induced subgraph, suggesting that
it is a good strategy to use when sorting the list of graphs.

@@

We now test whether each of our ordering strategies results, as hoped,
in a reduced number of calls to the subgraph isomorphism function
in \Cref{alg:brute-force}.  As our first test case, we consider 
$\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ ---
finding all induced universal graphs of order 9 for the family of
all graphs of order 5.  We consider four strategies.
The first two of these are the automorphism-count and `degree extremeness'
measures described above.  The fourth strategy is a random order.
Finally, we used one further strategy, which we call ``almost random''.
Under this strategy, $I_5$ and $K_5$ are placed at the start of the list,
and the remaining 32 graphs are in random order.  This strategy was not
intended to be useful in practice, but to give insight into whether
the success of the first two strategies was purely due to having
$I_5$ and $K_5$ at the start of the list of graphs.
For each of these strategies, we ran the program 50 times.

The results are shown in \Cref{fig:second-experiment}.  Over 50 runs,
the `automorphisms` strategy required a mean of 304,758 calls to the subgraph
isomorphism solver, while the `degree' strategy required a very similar
number of calls:  a mean of 304,701.  Surprisingly, the `almost random'
strategy had the lowest average number of calls of all the strategies,
at 304,431.  The `random' strategy fared poorly, requiring 822,649
calls on average and a minimum of 451,287 calls in the best of the 50 runs.
The variance in number of calls was tiny for all but the random strategy,
with between 304,000 and 305,000 calls being required for each run.

@@

    \caption{Using each of our four order strategies, we ran
        $\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ 50 times. The figure
        shows the number calls to the subgraph isomorphism algorithm, in thousands.
        Horizontal lines show mean number of calls.}
\label{fig:second-experiment}
\end{figure}

We note, first, that `degree' strategy appears to be at least as effective
as the `automorphisms' strategy for this instance, despite the lower
correlation observed in \Cref{fig:scatter}.  Perhaps most surprising is
that the `almost random' strategy performed better than any of the other
strategies; our expectation had been that this simple strategy would
give some but not all of the benefit of the first two strategies.
We conjecture that the almost-random strategy is effective because it
quickly checks a diverse range of graphs of graphs.  Consider, for example,
\Cref{tab:graphs-by-autom-count}, which shows the graphs of order 5 listed
in descending order of automorphism count.  This is the order in which they
are checked under the automorphisms strategy (with the order within each
row chosen randomly).  Note that the two graphs in the second row are
similar to the graphs in the first row; the first graphs in each of the first
two rows, for example, have a maximum common induced subgraph of size 4.
It seems likely that if a graph contains both graphs in the first row of
the table as induced subgraphs, then it may also contain the graphs in the
second row. The `almost random' strategy is likely to choose graphs
that are quite different from a clique and an independent set after the first
two graphs, perhaps increasing the probability that one of these graphs
will not be contained in the larger graph.

@@

\subsection{Testing the strategies: $\calF \subset \calF{5}$}

While the `almost random' strategy was very effective when searching
for induced universal graphs of \emph{all} graphs of a given order,
it seems likely that it will be less effective if $\calF$ is a strict
subset of $\calF(k)$, and in particular if it does not include one
or both of $I_k$ and $K_k$.  \Cref{fig:second-experiment-using-sample}
examines this claim.  We generated 25 (TODO 50) subsets of $\calF(5)$,
with each subset containing exactly half of the elements of $\calF(5)$.
For each of these, we ran $\FuncSty{AllInducedUniversalGraphs}(\calF,9)$
using each of the four strategies.  TODO say what happened

@@

    \caption{Using each of our four order strategies, we ran
        $\FuncSty{AllInducedUniversalGraphs}(\calF,9)$ 50 times, where
        $\calF$ is a random subset of $\calF(5)$ containing 17 graphs. The figure
        shows the number calls to the subgraph isomorphism algorithm, in thousands.
        Horizontal lines show mean number of calls.}
\label{fig:second-experiment-using-sample}
\end{figure}

>>>>

We write $\calF(k)$ to denote the family of all graphs on $k$ vertices,
and we write $f(k)$ to denote the order (that is, the
number of vertices) of a minimal induced universal graph for $\calF(k)$.
We write $F(k)$ to denote the number of non-isomorphic graphs of order $f(k)$
that are induced universal for $f(k)$.
To give an example, we have $f(3)=5$ and $F(3)=5$; all five of the minimal
induced universal graphs
for $\calF(3)$ are shown in \Cref{fig:univ3a}. \Cref{fig:univ3b} shows the four
graphs in $\calF(3)$ as induced subgraphs of a single induced universal graph.

@@

    \par\bigskip
    \subfigure[][A demonstration that the first graph in \Cref{fig:univ3a} is induced universal
            for $\calF(3)$.  For each graph $G$ in $\calF(3)$ ($I_3$, $K_3$, $P_3$, and
            a graph with a single edge), an induced subgraph isomorphic to $G$ is shown in a
            single color.] {
      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                clockwise=5] {
                    0,1[fill=uofgpumpkin],2[fill=uofgpumpkin],3[fill=uofgpumpkin],4;
           3 -- 0;
           4 -- 0;
           4 -- 1;
           4 -- 3;
        };
      \end{tikzpicture}
      \qquad
      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                clockwise=5] {
                    0[fill=uofgturquoise],1,2,3[fill=uofgturquoise],4[fill=uofgturquoise];
           4 -- 1;
           3 --[uofgturquoise, ultra thick] 0;
           4 --[uofgturquoise, ultra thick] 0;
           4 --[uofgturquoise, ultra thick] 3;
        };
      \end{tikzpicture}
      \qquad
      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                clockwise=5] {
                    0,1[fill=uofgrose],2,3[fill=uofgrose],4[fill=uofgrose];
           3 -- 0;
           4 -- 0;
           4 --[uofgrose, ultra thick] 1;
           4 --[uofgrose, ultra thick] 3;
        };
      \end{tikzpicture}
      \qquad
      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                clockwise=5] {
                    0,1[fill=uofgmocha],2[fill=uofgmocha],3,4[fill=uofgmocha];
           3 -- 0;
           4 -- 0;
           4 --[uofgmocha, ultra thick] 1;
           4 -- 3;
        };
      \end{tikzpicture}
      \label{fig:univ3b}
    }
\caption{Minimal induced universal graphs for the family of all graphs on 3 vertices}
\label{fig:univ3}
\end{figure}

This paper uses a brute-force approach similar to that of Preen to find minimal
induced universal graphs for families of small graphs.  Our contributions
are (1) an ordering strategy to reduce the number of subgraph isomorphism
calls required by the brute force algorithm; (2) exact values for $f(k)$ $(4 \leq k \leq 6)$
and $F(k)$ $(1 \leq k \leq 5)$; (3) an upper bound of 18 for $f(7)$; (4) a hill-climbing
search algorithm to find small (but not necessarily optimal) induced universal graphs
(5) exact results for families of small trees.

The sequel is structured as follows.
Section~\ref{sec:method} describes the basic brute force method.
Section~\ref{sec:iteration-order} compares four methods for sorting
the graphs in $\calF$, and compares their effect on the number of subgraph
isomorphism calls made by the brute-force algorithm.
Sections~\ref{sec:results5} presents exact
values of $f(k)$ and $F(k)$ for $k \leq 5$.
Section~\ref{sec:f6} gives the value of $f(6)$ and and
Section~\ref{sec:f7} gives bounds on $f(7)$; in each of these sections, the lower
bound is proven and a graph obtained by heuristic search demonstrating the
upper bound is given.  Finally, Section~\ref{sec:trees} gives results for
families of trees.

@@

Given $\calF$ and $n$,
we use the brute-force approach
shown in \Cref{alg:brute-force} --- which is essentially the same as the method
described by Preen \cite{preen_math_se} ---
to find the set of
$n$-vertex graphs that are induced universal for the family $\calF$.
The function $\FuncSty{AllInducedUniversalGraphs}()$ is the entry point.
\Lineref{TryEachG} tests in turn each candidate graph $G$ from the family
of $n$-vertex graphs, and adds to the collection $\calG$ those that are induced universal
for $\calF$.  The function $\FuncSty{IsInducedUniversal}()$ tests
whether a graph $G$ is induced universal for $\calF$ by checking for
each $H \in \calF$ that $H$ is isomorphic to a subgraph of $G$.

For the collection $\calF(n)$ of candidate graphs on \lineref{TryEachG}, our
implementation uses graphs generated using Brendan McKay's \texttt{geng}
program \cite{DBLP:journals/jal/McKay98}
\footnote{The program \texttt{geng} is distributed
with Nauty; we used version 27.r3.\url{https://pallini.di.uniroma1.it/}}.
These are read in \texttt{graph6}
format\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/formats.html}}
from a text file as the program proceeds, and therefore only one candidate
graph at a time needs to be stored in memory.

Our implementation is written purely in Python and run using the CPython
interpreter.
On line \lineref{CallSubIso}, the function $\FuncSty{InducedSubIso}()$
calls an algorithm for the induced
subgraph isomorphism decision problem; for this,
our program uses an implementation by the author of the McSplit
algorithm \cite{DBLP:conf/ijcai/McCreeshPT17}.
McSplit was designed for
the more general maximum common induced subgraph problem,
but it can be used for induced subgraph isomorphism with a simple
modification: we backtrack when the calculated upper bound
on the order of a common subgraph is less the order of either
of the two input graphs.
This method for solving
the subgraph isomorphism problem is very fast if both input graphs
are small, as the graphs considered in this paper are.
While McSplit is suitable for our purposes in this paper,
we do not claim that it is the fastest induced subgraph isomorphism
solver for very small graphs; it would be useful future work to perform
an experimental comparison with other subgraph isomorphism solvers.
%%% (Although
%%% I have described the application of this method to families of all graphs of a given order,
%%% the method could easily be applied to any family $\calF$ of graphs, and indeed
%%% Section~\ref{sec:trees} gives results from the application of this method
%%% to families of trees.)

\begin{algorithm}[h!]
\DontPrintSemicolon
\nl $\FuncSty{IsInducedUniversal}(\calF,G)$ \;
\nl \KwData{A family of graphs $\calF$ and a graph $G$.}
\nl \KwResult{A logical value indicating whether $G$ is induced universal for $\calF$.}
\nl \Begin{
\nl    \For {$H \in \calF$ \label{HLoop}}{
\nl      \lIf {$\neg\FuncSty{InducedSubIso}(H, G)$\label{CallSubIso}}{$\KwSty{return}$~$\AlgVar{false}$}
       }
\nl   $\KwSty{return}$~$\AlgVar{true}$ \;
    }
\medskip
\nl $\FuncSty{AllInducedUniversalGraphs}(\calF,n)$ \label{BruteForceFun} \;
\nl \KwData{A family of graphs $\calF$ and a natural number $n$.}
\nl \KwResult{The set of all graphs of order $n$ that are induced universal for $\calF$.}
\nl \Begin{
\nl    $\calG \gets \emptyset$ \;
\nl    \For {$G \in \calF(n)$}{
\nl      \lIf {$\FuncSty{IsInducedUniversal}(\calF,G)$\label{TryEachG}}{$\calG \gets \calG \cup \{G\}$}
       }
\nl   $\KwSty{return}$~$\calG$ \;
    }
\caption{A brute-force algorithm for finding all order-$n$ induced universal subgraphs of a family $\calF$ of graphs.
	The entry point is $\FuncSty{AllInducedUniversalGraphs}(\calF,n)$.}
\label{alg:brute-force}
\end{algorithm}

The full set of experiments described in this paper, run sequentially,
took less than a day to complete on a laptop with an Intel Core i5-6200U CPU
and 8 GB RAM.\footnote{The code and
results from this paper, including lists of minimal induced universal graphs
in graph6 format, are available from
\url{https://github.com/jamestrimble/small-universal-graphs}}

@@

When deciding on \lineref{HLoop} of the algorithm whether a graph $G$ contains
every element of $\calF$ as an induced subgraph, are some iteration orders of
$\calF$ better than others?  For a graph $G$ that contains a copy of each
element of $\calF$, the iteration order is of no importance;
every element of $\calF$ must be checked before $\AlgVar{true}$ is returned.
For a graph $G$ that
does \emph{not} contain a copy of each element of $\calF$, however, we would like
to iterate over $\calF$ in an order such that graphs that are likely to fail the
subgraph isomorphism test are checked early; that way, we can quickly return
$\AlgVar{false}$ and avoid unnecessary calls to the subgraph isomorphism algorithm.

We begin by considering the case where our family of graphs $\calF$
is $\calF(k)$ for some $k$; that is,
$\calF$ is the set of all graphs of order $k$.
As Diaconis and Chatterjee note, \cite{chatterjee2021isomorphisms}, the complete
graph $K_n$ is the `hardest' graph of order $n$ for a random graph to contain.
This led me
to consider two orderings in which $K_k$ and its complement $I_k$ come
first.  The first of these approaches is to the sort $\calF(k)$ such that
graphs with large automorphism groups appear first.  (Intuitively, if a graph
has few automorphisms then we can generate many different labelled graphs
by permuting the vertex labels.  This gives more `opportunities' for the
graph to be an induced subgraph of $G$.)
The second approach considered is to place graphs
with unusually high or low edge counts, as measured by the absolute value of
${2|E(G)| - {|V(G)| \choose 2}}$, first.

\Cref{fig:scatter} examines whether graphs that appear early in the orderings
of $\calF$
generated by each of these two strategies are contained in few graphs with
a given number of vertices, as we would hope to be the case.
For this experiment, we let $\calF = \calF(5)$, and we consider the problem
of finding induced universal graphs of order 8.
Each point on the plots
represents one of the 34 graphs of order 5 (fewer than 34 points are visible
due to exactly-overlapping points, as we would expect).  On the vertical
axis, we have the number of graphs in $\calF(8)$ containing $G$; we would like
graphs for which this value is small to appear early in our order.
On the horizontal axes we have each of our two measures: size of automorphism group
in the first plot and `extremeness' of edge count in the second plot.
As we can see from the first plot, the automorphism measure has a strong correlation
with the number of graphs containing $G$ as an induced subgraph, suggesting that
sorting by this measure could be a good strategy.

@@

We now test whether each of our ordering strategies results, as hoped,
in a reduced number of calls to the subgraph isomorphism function
in \Cref{alg:brute-force}.  As our first test case, we consider 
$\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ ---
finding all induced universal graphs of order 9 for the family of
all graphs of order 5.  We consider four strategies:

The first two of these are the automorphism-count and `edge-count extremeness'
measures described above.  The fourth strategy is a random order.
Finally, we used one further strategy, which we call ``almost random''.
Under this strategy, $I_5$ and $K_5$ are placed at the start of the list,
and the remaining 32 graphs are in random order.  This strategy was not
intended to be useful in practice, but to give insight into whether
the success of the first two strategies was purely due to having
$I_5$ and $K_5$ at the start of the list of graphs. \Cref{tab:ordering-strategies}
summarises these ordering strategies.

\begin{table}[h!]
\centering
 \begin{tabular}{p{0.25\linewidth} p{0.55\linewidth}}
 \toprule
    Name & Description of strategy \\ [0.5ex]
 \midrule
    Automorphisms & Sort by size of the automorphism group \\
    Edges & Sort by number of edges \\
    Almost-random & Shuffle the list of graphs in $\calF$, then move $K_k$ and $I_k$ to the start (in random order) \\
    Random & Shuffle the list of graphs in $\calF$ \\
 \bottomrule
\end{tabular}
\caption{Ordering strategies tested for the brute-force algorithm}
\label{tab:ordering-strategies}
\end{table}

For each of these strategies, we ran the program 50 times.

The results are shown in \Cref{fig:second-experiment}.  Over 50 runs,
the automorphisms strategy required a mean of 304,746 calls to the subgraph
isomorphism solver, while the edges strategy required marginally fewer
calls:  a mean of 304,701.  Surprisingly, the almost-random
strategy had the lowest average number of calls of all the strategies,
at 304,387.  The random strategy fared poorly, requiring over 800,000
calls on average.
The variance in number of calls was small for all but the random strategy,
with between 304,000 and 305,000 calls being required for each run.

\begin{figure}[h!]
    \centering

@@

    \caption{Using each of our four order strategies, we ran
        $\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ 50 times. The plot
        shows the number calls to the subgraph isomorphism algorithm, in thousands.
        Horizontal lines show average (mean) number of calls.}
\label{fig:second-experiment}
\end{figure}

We note, first, that edges strategy appears to be at least as effective
as the automorphisms strategy for this instance, despite the lower
correlation observed in \Cref{fig:scatter}.  Perhaps most surprising is
that the almost-random strategy performed better than either of these two
strategies; our expectation had been that this simple strategy would
give some but not all of the benefit of the first two strategies.
We conjecture that the almost-random strategy is effective because it
quickly checks a diverse range of graphs.  Consider, for example,
\Cref{tab:graphs-by-autom-count}, which shows the graphs of order 5 listed
in descending order of automorphism count.  This is the order in which they
are checked under the automorphisms strategy (with the order within each
row chosen randomly).  Note that the two graphs in the second row are
similar to the graphs in the first row; the first graphs in each of the first
two rows, for example, have a maximum common induced subgraph of size 4.
It seems likely that if a graph contains both graphs in the first row of
the table as induced subgraphs, then it may also contain the graphs in the
second row. The almost-random strategy is likely to choose graphs
that are quite different from a clique and an independent set after the first
two graphs, perhaps increasing the probability that one of these graphs
will not be contained in the larger graph.

@@

\subsection{Testing the strategies: $\calF \subset \calF(5)$}

While the almost-random strategy was very effective when searching
for induced universal graphs of \emph{all} graphs of a given order,
it is plausible that it will be less effective if $\calF$ is a strict
subset of $\calF(k)$, and in particular if it includes
neither $I_k$ nor $K_k$.  \Cref{fig:second-experiment-using-sample}
examines this claim.  We generated 50 subsets of $\calF(5)$,
with each subset containing exactly half of the elements of $\calF(5)$.
For each of these, we ran $\FuncSty{AllInducedUniversalGraphs}(\calF,9)$
using each of the four strategies.  The `automorphisms' strategy
was the best overall in this case, requiring on average $8\%$, $27\%$, and $52\%$
fewer calls to the subgraph isomorphism function than the edges, almost-random,
and random strategies respectively.

To help us understand why the almost-random
strategy performs poorly for these instances, we connect the four solver runs on
each instance together on the plot with blue lines.  The almost-random
strategy, on the instances where it performs poorly, requires
exactly the same number of subgraph isomorphism calls as the random strategy.
These are the instances in which $\calF$ contains neither $K_5$ nor $I_5$;
thus, the almost-random and random strategies are equivalent.  Notably,
for these instances it is also the case that the automorphisms strategy
tends to outperform the edges strategy.

@@

    \caption{Using each of our four order strategies, we ran
        $\FuncSty{AllInducedUniversalGraphs}(\calF,9)$ 50 times, where
        $\calF$ is a random subset of $\calF(5)$ containing 17 graphs. This
        parallel coordinates plot
        shows the number calls to the subgraph isomorphism algorithm, in thousands.
        Horizontal black lines show average (mean) number of calls.  The four runs for
        each instance are connected by blue lines.}
\label{fig:second-experiment-using-sample}
\end{figure}

\subsection{Testing the strategies: a family of trees}

As a final test case, we let $\calF$ be the family of all trees on 6 vertices,
and searched for all graphs of order 8 that are induced universal for $\calF$.
\Cref{fig:second-experiment-using-trees} shows the results. The edges, almost-random,
and random strategies had identical results, since all members of $\calF$ have five
edges and the family contains neither $K_6$ nor $I_6$.  The automorphisms strategy
requires fewer subgraph isomorphism calls on average than the other three strategies,
although on seven of the 50 instances it does require marginally more calls than
the other strategies.

    \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot-using-trees}

    \caption{Using each of our four order strategies, we ran
        $\FuncSty{AllInducedUniversalGraphs}(\calF,8)$ 50 times, where
        $\calF$ is the set of all trees of order 6. The plot
        shows the number calls to the subgraph isomorphism algorithm, in thousands.
        Horizontal black lines show average (mean) number of calls.}
\label{fig:second-experiment-using-trees}
\end{figure}

The `automorphisms' strategy performs consistently well across our three test
cases; therefore, we use this strategy for the remainder of the paper.
